# Workshop Name: Subtitle

The beginning of every workshop notebook should be an H1 (one hashtag) header, with the workshop name, followed by a colon, followed by a subtitle for this notebook.

Provide a brief introduction to what this lesson covers here.

## Workshop Subtopic 1

Provide some more text on what this subtopic covers.

```{r}
# You might want to have an install cell, if your workshop uses a special package
# e.g.,
# install.packages(PACKAGE)
```

```{r}
# You might want to have an import cell, so that you can do all imports at the beginning.
# e.g.,
# library(PACKAGE)
```

Be sure to have text between code cells explaining the material, step-by-step.

### Workshop Sub-subtopic 1

Perhaps your subtopic will be divided into finer sub-subtopics. Be sure to use consistent headers for each one, and always have an introductory paragraph explaining what the workshop attendee can expect with this section.

### Challenge #: Challenge Title Here

Every challenge should be an H3 (3 hashtags) header. The challenge should be numbered, and should have a title.

Insert some text here explaining the challenge.

```{r}

# You may have some starter code for the challenge that you can put in its own cell
# Always have a following cell that says "YOUR CODE HERE", so that attendees know where to put their code
```

```{r challenge_X}
# YOUR CODE HERE
```

# Optional: Closing Section

You may optionally include a summary section at the end of the workshop highlighting what this notebook covered.