---
title: "R Fundamentals - Part 3 of 4: Data Frames"
output: html_notebook
---

## Learning Objectives

Welcome to Part 4 of the R Fundamentals series. Our goals for today's workshop are:

1.  Think through ways to visualize our data

2.  Use `ggplot` to create a boxplot and scatter plot

3.  Learn about custom functions

Throughout this workshop series, we will use the following icons:

🔔 **Question**: A quick question to help you understand what's going on.

🥊 **Challenge**: Interactive exercise. We'll go through these in the workshop!

⚠️ **Warning**: Heads-up about tricky stuff or common mistakes.

💡 **Tip**: How to do something a bit more efficiently or effectively.

🎬 **Demo**: Showing off something more advanced so you know what you can use R for in the future

Before we get started, lets load in the `tidyverse` package that we used in Part 2. We will use this today for subsetting our data and for making plots.

```{r}
# if you haven't install tidyverse before, uncomment this line 
#install.packages("tidyverse")

# if you have previously installed tidyverse, load it into the current R session
library(tidyverse)
```

### 🥊Challenge 1: Reading in data

Load in `data/gapminder.csv` using the `read.csv()` function from earlier workshops and assign it to the object **gap**.

```{r}



```

## Visualizing Data

In Part 3 we used `summary()`, `table()`, and `mean()` to learn some things from our data. For example:

### 🥊Challenge 2: What is the average life expectancy in our data? How much does life expectancy range?

```{r}

mean(...)

summary(...)

```

We can combine our summary statistics functions with the subsetting functions `filter()` and `select()` from `tidyverse` to ask more specific questions. **For example, what is the average life expectancy in Asia?**

```{r}

gap_asia <- gap %>%
  filter(...)

summary(gap_asia$lifeExp)

```

A great way to ask more involved questions about our data and convey this information is to make a plot!

**🎬 How does the life expectancy compare across continents? A box plot example...**

```{r}

gap %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot(aes(fill = continent)) +
  theme_bw() +
  ylab("Life Expectancy (years)") +
  ggtitle("Life expectancy by continent (1952-2007)")

  
```

**🔔 What are some things we learn from this plot?**

### ggplot

`ggplot` is a package from `tidyverse` that helps create highly customizable plots, all of which have the same syntax underlying them.

A `ggplot` is comprised of three fundamental building blocks:

1.  **data**: typically as a data.frame

2.  **"aes"thetics**: how our variables will be visually represented (x and y axes, colors, shapes etc.)

3.  **geoms**: what type of plot will represent our data (histogram, scatter plot, bar chart etc.)

One thing to remember is that `ggplot` works in layers, similar to photoimaging software such as Photoshop. We create a base layer, and then stack layers on top of that base layer. New layers are added to a plot using the addition symbol `+`. Each of the building blocks above is added as a new layer.

## Box plots

Here we re-create the boxplot from above.

1.  **data**: the first layer of a `ggplot` is always the data. This creates a blank canvas on which to add other layers. For us, the data is the `gap` data.frame.

```{r}

ggplot(data = gap)

```

2.  **"aes"thetics**: next we specify which variables we want to visualize and how. Here we want `continent` on the x-axis, `lifeExp` on the y-axis, and `continent` also as the fill color. All aesthetics go inside the `aes()` function. This prepares our blank canvas.

```{r}

ggplot(data = gap, aes(x = continent, y = lifeExp, fill = continent))
```

3.  **geom**: finally we specify what type of plot we actually want to visualize our data. For us, this is a box plot. Note how we add a new layer with the `+` (not the pipe `%>%` or `|>`). Now our canvas has a picture!

```{r}
ggplot(data = gap, aes(x = continent, y = lifeExp, fill = continent)) +
  geom_boxplot()
```

Any other things that we want to add or customize about our plot we also add as layers. To recreate the plot above, we can add a title, change the axis titles, and change the theme (the overall look of the plot).

```{r}

ggplot(data = gap, aes(x = continent, y = lifeExp, fill = continent)) +
  geom_boxplot() +
  # add a title
  ggtitle("Life expectancy by continent (1952-2007)") +
  # change the y axis title
  ylab("Life Expectancy (years)") +
  # change the x axis title
  xlab("Continent") +
  # change the overall theme 
  theme_bw() +
  # remove the legend, b/c this is redundant
  theme(legend.position = 'none')
  

```

### 🥊Challenge 3: What plots could we create from our data?

Explore the variables in the `gap` data frame. What other questions could we ask from our data? What plot would answer the question?

-   Q: Which continent has the most varied life expectancy? A: Box plot showing distribution of life expectancy by continent.

-   Q: ... A: ...

💡 A helpful way to learn how to create many different types of plots is to look at a [gallery](https://r-graph-gallery.com/index.html) of possible plots in R.

## Scatter plots

Box plots are good for looking at the distribution of a single continuous variable (eg. life expectancy) and showing that for different groups (eg. continents). To show the relationship between two continuous variables a scatter plot is better. For example, **has life expectancy increased with time in Oceania?**

To make a scatter plot, we use the same syntax as above, but rather than `geom_boxplot()` we use `geom_point()`.

```{r}

gap %>%
  # limit data to just counties in oceania
  filter(continent == "Oceania") %>%
  ggplot(aes(x = year, y = lifeExp)) +
  geom_point()

```

**🔔 Why are there two points in each year?**

```{r}
# we can color the points to indicate which country they represent - this is another aesthetic 

gap %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_point()
```

**🔔 What can we add to the plot to make it more informative?**

```{r}

gap %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_point() +
  # add more layers to make the plot more informative
  ...

```

### 🥊Challenge 4: Create a scatter plot

Create a scatter plot that shows the relationship between `lifeExp` and `gdpPercap` . Start with a "simple" plot using `geom_point()`. Then think about edits you could make to the plot that might make it easier to read. For example, changing the x and y axis, adding a title, coloring the points by a certain grouping etc.

```{r}

gap %>%
  ggplot(...)

```

💡`ggplot` can be used to create many different plots ([here](https://r-graph-gallery.com/index.html) are some examples with the underlying code). To learn more about using `ggplot` you can take D-Lab's [R Data Visualization](https://github.com/dlab-berkeley/R-Data-Visualization).

## Custom Functions

Often times, we want to do the same thing over and over again in our code but with minor changes. For example, we might want to plot the life expectancy against GDP per capita for multiple continent in our data.

```{r}

# plot gdppercap vs. lifeExp for Oceania 
gap %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  ggtitle("Life expectancy against GDP per capita in Oceania") +
  ylab("Life expectancy (years)") +
  xlab("GDP per capita (dollars)") +
  theme_bw()
```

If we wanted to show this for the Americas, we could copy and paste the code and change the continent.

```{r}
# plot gdppercap vs. lifeExp for the Americas 
gap %>%
  filter(continent == "Americas") %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point() +
  ggtitle("Life expectancy against GDP per capita in Oceania") +
  ylab("Life expectancy (years)") +
  xlab("GDP per capita (dollars)") +
  theme_bw()
```

**🔔 What is wrong with the plot above? Have we made a mistake?**

Instead of copying and pasting code, we can instead put the code into a **function** and call the function whenever we want to create the plot for different continents.

We have used many functions that are part of base R or `tidyverse`, for example `head()`, `length()`, `summary()` etc. Custom functions are the same but we define the name of the function, what the input is, and what the output is.

The structure of a custom function is:

```{r}
# # structure of code
# function_name <- function(optional arguments){
#   
#   code
# 
# }

# example
print_number <- function(x){
  
  paste0("My number is: ", x)
  
}

print_number(1)
print_number(5)

```

Custom functions in R have four main components:

1.  The `function` keyword: this keyword tells R that we are defining a function

2.  The **arguments**: these are the inputs to the function that change. In the example, our argument was a variable `x` that we can make any number we want. Functions can have multiple arguments and they can be any class of object.

3.  The **code**: this is the code that the function executes each time it is called. In our example, we just have a single line of code pasting together the sentence "My number is" with the argument `x`. The code in a function can be as long as needed.

4.  The **output**: by default, a function returns the last thing that was outputted by the code. It is good practice to add a `return()` statement, in which you put the explicit thing you want outputted.

💡`paste0` concatenates things together into a single string/sentence. It is a helpful function to use inside custom functions.

Here is another example for a custom function that returns the average life expectancy of each continent.

```{r}
continent_lifeexp <- function(cont){
  
  # it is good practice to comment what the function does and what it's arguments are at the top:
  # function to return life expectancy of continent
  # cont = name of continent, string
  
  # everytime we would have written the continent, we now write 'cont'
  # limit to continent cont
  gap_subset <- gap %>%
    filter(continent == cont)

  # save the mean life expectancy for that continent
  lifeexp_mean <- mean(gap_subset$lifeExp)
  
  # return the life expectancy
  return(paste0("The average life expectancy in ", cont, " is ", lifeexp_mean))
    
}

```

Running this chunk of code does not run the code inside the function. It saves a function called `continent_lifeexp()`, which now appears in our global environment and can be called at any time in our session.

```{r}
# run the function for different continents
continent_lifeexp("Asia")
continent_lifeexp("Americas")
```

### 🥊Challenge 5: Write a custom function

Write a function that filters the gap data set to a specific year. Then estimate the average GDP per capita in that year. Make your function so you can input a different year and the functions outputs the average GDP per capita for that year. Use the code above as a template.

💡When writing functions, it is helpful to make sure the inside code does what you want it to without error before including it in the function. To do this, write the code as you would not in a function, check it works, then replace any year values with the arugments that you use for year.

```{r}






```

### Functions with ggplot

Returning to our example, we can create a function that takes the name of the continent as an argument and outputs plot of life expectancy against GDP for each continent.

```{r}

gdp_lifeexp_plot <- function(cont){
  
  # function plot life exectancy against GDP per capita by continent 
  # cont = name of continent, string

  # everytime we would have written the continent, we now write cont 
  p <- gap %>%
    filter(continent == cont) %>%
    ggplot(aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    # here we use the paste0() function to allow cont to change in the title
    ggtitle(paste0("Life expectancy against GDP per capita in ", cont)) +
    ylab("Life expectancy (years)") +
    xlab("GDP per capita (dollars)") +
    theme_bw()
  
  # return the plot so it prints out when we call the function
  return(p)
  
}

```

Now we have a function called `gdp_lifexp_plot()` in our global environment. We can call it at any time and provide different continent names.

```{r}

gdp_lifeexp_plot("Africa")
gdp_lifeexp_plot("Europe")

```

Functions are a great tool for improving your code, especially when you find yourself copying and pasting code. To read more about functions we recommend the [Functions chapter in R for Data Science](https://r4ds.hadley.nz/functions.html).

## Key Points

-   Using plots to visualize data is a great way to convey information and find patterns in your data

-   `ggplot` is a great for creating plots

-   Custom functions are used to make your code easier to write and less error prone - anytime you are copying and pasting chunks of code, consider putting it in a custom function and calling that function when you need to

## Next steps...

Congratulations on finishing D-Lab's R Fundamentals Series. If you would like to continue learning more R, we recommend taking the following workshops, which build on the skills learned in this workshop series.

1.  [R Data Wrangling](https://github.com/dlab-berkeley/R-Data-Wrangling) - learn more `tidyverse()` commands, including how to create summary statistics by groups and to re-shape data

2.  [R Data Visualization](https://github.com/dlab-berkeley/R-Data-Visualization) - learn more `ggplot()` tools and create a wide array of different data visualizations
