---
title: "R Fundamentals - Part 3 of 4: Data Frames"
output: html_notebook
---

## Learning Objectives

Welcome to Part 3 of the R Fundamentals series. Our goals for today's workshop are:

1.  Use conditional statements to filter rows.

2.  Practice generating summary statistics.

3.  Learn how to store categorical variables as factors.

Throughout this workshop series, we will use the following icons:

🔔 **Question**: A quick question to help you understand what's going on.

🥊 **Challenge**: Interactive exercise. We'll go through these in the workshop!

⚠️ **Warning**: Heads-up about tricky stuff or common mistakes.

💡 **Tip**: How to do something a bit more efficiently or effectively.

📝 **Poll**: A zoom poll to help you learn.

🎬 **Demo**: Showing off something more advanced so you know what you can use R for in the future


## Tidyverse

In Part 2, we learned how to subset columns in **base R**. Base R refers to all the functions and operators that you can use in R without installing any additional packages. Base R will get you a long way in data science. However, it is not always the easiest code to read and interpret. 

A common package that people use to perform data science in R is `tidyverse`. The `tidyverse` package includes functions that are designed to be more intuitive for doing data cleaning, visualization, and analysis.

Before we use functions from `tidyverse`, we need to take two steps:

1.  **Install** the package - you only need to do this once on your computer.

2.  **Load** the package into R - you need to do this every time you open a new R session.

💡 **Tip:** These steps are the same for every package you download in R.

```{r}
# install the tidyverse package - note the name of the package needs to be in quotes
install.packages("tidyverse")
```

⚠️ **Warning:** If you're running into issues with installing `tidyverse` the best thing to do is switch to DataHub (link in the README on [github](https://github.com/dlab-berkeley/R-Fundamentals)). All the material and packages are pre-loaded there.

```{r}
# load the package into the current R session - note the name of the package doesn't need to be in quotes (!) and you can use tab complete to see a list of installed packages
library(tidyverse)

# check that tidyverse has been loaded
?tidyverse
```

⚠️ **Warning:** It's okay for text to display in red when you load in a package. **Warnings** are fine and don't mean anything is wrong. **Errors** are problems and mean something is wrong that needs to be fixed.

### Subsetting Columns in Tidyverse

Now we have installed and loaded tidyverse, we can use its `select` function to subset variables. First, we need our dataframe.

### 🥊 Challenge 1: Reading in Data

In Part 2 we learned to read in a .csv file and assign it to a data frame called `gap`. Let's do that again. How can you find out what variables we have in our data?

```{r}




```

Next we can look up the help files for the `select()` function from tidyverse.

```{r}

# look up the documentation for select - focus on the Arguments section
?select

# select year, pop, and lifeExp from the dataset 
select(gap, year, pop, lifeExp)

# assign these variables to a new dataset
gap_subset <- select(gap, year, pop, lifeExp)
dim(gap_subset)
head(gap_subset)

```

💡 **Tip:** If you want to learn how to use tidyverse in more detail, you can check out D-Lab's [R Data Wrangling and Manipulation](https://github.com/dlab-berkeley/R-Data-Wrangling) workshop series.

### 🥊 Challenge 2: Select Columns using Tidyverse

Use the `select()` function from `tidyverse` to select three variables from the gap data frame. Assign those three variables to a new data frame. Check the dimensions of your new data frame using the `dim()` function.

```{r}




```

## Filtering Data Frames

If we open the `gap` data frame (click on the gap object in the Global Environment panel), we see a **Filter** button. This allows us to subset the data frame by rows. We can specify that we want to only show rows for Pakistan or only show rows in the year 1992. We can achieve the same thing by writing code.

### Conditional Statements

In order to specify which rows we want to subset, we need to write **conditional statements** that specify which rows we want. Conditional statements check whether something is true or false. To write them, we need to use **logical operators**:

-   `>` bigger than

-   `<` less than

-   `==` equals to (note the double equal signs)

-   `!=` not equal to

-   `%in%` is in

📝 **Poll RFun 3-1:** Are the following statements TRUE or FALSE? Answer in the Zoom poll.

```{r}
# check whether the average population is bigger than 5 million 
mean(gap$pop) > 5000000

# check whether the minimum year is less than or equal to 1950?
min(gap$year) <= 1950

# check if there are exactly 6 variables in the gap data
length(names(gap)) == 6
```

### Filtering

We can use the same logical operators to filter (subset) the rows in our data. We will use the function `filter` from `tidyverse`.

```{r}
# look up the documentation for filter
?filter

# return all rows for Pakistan
filter(gap, country == "Pakistan")

# return all rows where the life expectancy is bigger than 80
filter(gap, lifeExp > 80)
```

We can also combine conditional statements using the `&` and `|` (or) operators. `&` checks both statements and returns rows that meet **both** conditions. `|` (or) checks both statements and returns rows that meet **either** conditions.

```{r}
# return all rows in Asia with a GDP per capita above or equal to 30K
filter(gap, continent == "Asia" & gdpPercap >= 30000)

# return all rows in Germany or France
filter(gap, country == "Germany" | country == "France")

# return all rows in Mexico but not from the year 2007
filter(gap, country == "Mexico" & year != 2007)
```

### Piping

We can combine our `select` and `filter` commands to subset both columns and rows. An intuitive way to write code that combines multiple functions is to write multiple lines of code that feed into each other. We do this with the pipe operator `%>%` from `tidyverse`. The `%>%` operator takes the output on the left-hand side and passes it into the function on the right-hand side.

```{r}
# start with the gap data set
gap %>%
  # filter to Mexico - note, we don't need to specify gap inside select() any more
  filter(country == "Mexico") %>%
  # select the columns we want to return
  select(country, year, pop, lifeExp) 

```

We can assign the subset of our original data frame to a new data frame.

```{r}
# save Mexico rows to another data frame
data_mexico <- gap %>%
  # filter to Mexico
  filter(country == "Mexico") %>%
  # select the columns we want to return
  select(country, year, pop, lifeExp) 

dim(data_mexico)
```

💡 **Tip:** If you have R 4.1 or higher, there is now a base R pipe operator `|>` , which works the same as `%>%` but does not need `tidyverse` to have been loaded.

### 🥊 Challenge 3: Combining Filter and Select

Start with the gap data frame. Use the `%>%` operator and the functions `select()` and `filter`. Subset the data to countries in Asia or Oceania that have a life expectancy above 60, and select the country, year, life expectancy, and GDP per capita variables. Assign your subsetted data to a new data frame called **gap_subset**.

```{r}



```

## Missing Values

Before creating visualizations, computing basic summary statistics is a good first step after familiarizing yourself with your data. The `summary()` function gives statistics about the distribution of a variable.

```{r}
# summary for life expectancy
summary(gap$lifeExp)

# we can also print only the mean value using the mean() function
mean(gap$lifeExp)

```

🔔 **Question:** What do you notice when you use the `mean()` and `summary()` functions on `gap$gdpPercap` ?

```{r}
# try out mean() and summary() on gdpPercap

```

When we have missing values, functions may perform differently than expected. It is important to know if and why there are missing values in our data. In R, missing values are represented with `NA`, which stands for "not available" or "not applicable". We can use the `is.na()` function to count how many missing values there are in a variable.

```{r}
# count the number of missing values in gdpPercap
is.na(gap$gdpPercap) %>% 
  sum()

# we can also get this information from summary()
summary(gap$gdpPercap)

```

When we ran `mean(gap$gdpPercap)`, R returned `NA` because it can't calculate a mean over missing values. We need to tell it to explicitly ignore the missing values and calculate the mean using the non-missing values.

📝 **Poll RFun 3-2:** Look up the documentation for `mean`. Which argument will let us ignore `NA` values? What value should we set this argument to? Answer in the Zoom poll.

```{r}
# look up the documentation for mean
?mean

# calculate the mean of gdpPercap ignoring missing values 
mean(gap$gdpPercap, ...) # <- edit this

# is this the same as using summary?
summary(gap$gdpPercap)
```

⚠️ **Warning:** Some functions ignore `NA` values by default and some do not. The function documentation will tell you what the default behavior is, or you can find out with trial and error.

### 🥊 Challenge 4: Filtering Missing Values

We might want to drop the missing values from our data entirely. We can combine `filter()` and `is.na()` to remove rows that contain missing values. Edit the code below to drop all rows for which gdpPercapita is missing.

Reminder: `filter()` takes a conditional statement, checks if it's true, and then **keeps** all rows that the statement is true for. We want to keep all rows where there are **no** missing values for `gdpPercap`. You can check for missing with `is.na()`.

```{r}
# fill in a conditional statement in filter to keep only rows with non-missing values of gdpPercap
gap_nomissing <- gap %>%
  filter(...)

# the code below subtracts the number of rows in gap_nonmiss from number of rows in gap - this is how many rows filter() dropped
nrow(gap)-nrow(gap_nomissing)

```

```{r}

# check if we have any missing values for gdpPercap
summary(gap_nomissing$gdpPercap)

```

⚠️ **Warning:** When dropping missing values, R drops the entire row of data that contained those missing values. This means we lose cells that had non-missing values for other variables. Choosing whether to drop the row entirely or using functions that ignore missing values will depend on what you're trying to do in your data cleaning and analysis.

### 🎬 **Demo:**Reporting Missing Values for all Variables**

The code below uses two functions from **tidyverse**, `summarize()` and `pivot_longer()`, to make a table showing the number of missing values in each variable. We learn these functions in D-Lab's [R Data Wrangling](https://github.com/dlab-berkeley/R-Data-Wrangling) workshop, which is a great course to take after R Fundamentals.

```{r}
gap %>%
  summarize(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = 'variable', 
               values_to = 'NA_count')
```

## Categorical Variables

The functions `summary()` and `mean()` are helpful for **numeric** variables.

```{r}
# two types of numeric variables:

# (1) numeric
class(gap$lifeExp)

# (2) integer 
class(gap$year)
```

What if we try to use them on **character** variables?

```{r}
summary(gap$continent)

summary(gap$country)

```

Instead, we can use the `table()` function to print out a table of all the values that `continent` takes and how many times each value appears (i.e., its frequency).

```{r}
table(gap$continent)
```

We can also create a cross-table where we provide two variables to `table()`. It shows the counts for each cross of the two variables.

```{r}
table(gap$continent, gap$year)
```

### Factors

`country` and `continent` are both **categorical** variables, which means they have a fixed and known set of possible values. For example, `continent` can be one of five values: Africa, Americas, Asia, Europe, or Oceania.

⚠️ **Warning:** Not all character variables are categorical: a variable of people's names does not have a limited and known set of values, but it is a character variable. Likewise, not all categorical variables are characters: shoe size is a numeric variable but it has a fixed set of values, so it is a categorical variable.

In R, **categorical** variables are best stored as **factors** because this tells R that the variable can take one of a fixed set of values. This is helpful when performing calculations or making plots. The set of values that a factor variable can take are called **levels**. For the `continent` variable, the levels are Africa, Americas, Asia, Europe, and Oceania.

To convert a variable to a categorical variable, you use the `factor()` function. By default, it will look for all unique values in the variable and set those as the possible levels.

```{r}
# convert gap$continent into a factor variable
continent_factor <- factor(gap$continent)

# now we can use summary() on it
summary(continent_factor)

```

Factors are useful when it comes to making plots and tables, but we won't spend much time on them now.

💡 **Tip:** We created factor versions of our `continent` variable outside of the gapminder data frame. To create new variable or modify existing variables in a data frame, we can use the `mutate()` function from **tidyverse**. This is beyond this workshop but is covered in [R Data Wrangling](https://github.com/dlab-berkeley/R-Data-Wrangling).

## 🎬 Next time...

Selecting and filtering data are frames are key steps in telling stories from our data. Looking at summary statistics and visualization are the next steps for understanding our data and presenting it to other people.

🔔 **Question:** Run the chunk of code below. What do you think we are doing?

```{r}
gap %>%
  filter(year == 2007 | year == 1952) %>%
  select(gdpPercap, lifeExp, continent, year) %>%
  ggplot(aes(x = lifeExp, y = log(gdpPercap), color = continent, shape = continent)) +
  geom_point() +
  facet_wrap(~year) +
  ggtitle("Relationship between Life Expectancy and GDP per capita") +
  ylab("GDP per capita (logged)") +
  xlab("Life Expectancy (years)") +
  theme_bw() 

```

## Key Points

-   Use conditional statements with logical operators to filter data frames to a subset of rows.
-   The pipe operator `%>%` is a way to pass output from one line of code to the next line of code.
-   Different functions in R treat missing values differently - always be aware of whether you have missing data.
-   We used the functions `select()`, `min()`, `mean()`, `head()`, `filter()`, `summary()`, `is.na()`, `table()` and `factor()`.
