---
title: "R Fundamentals - Part 4 of 4: Visualization and functions"
output: html_notebook
---

## Learning Objectives

Welcome to Part 4 of the R Fundamentals series. Our goals for today's workshop are:

1.  Think through ways to visualize our data.

2.  Use `ggplot` to create a boxplot and scatter plot.

3.  Learn about custom functions.

Throughout this workshop series, we will use the following icons:

🔔 **Question**: A quick question to help you understand what's going on.

🥊 **Challenge**: Interactive exercise. We'll go through these in the workshop!

⚠️ **Warning**: Heads-up about tricky stuff or common mistakes.

💡 **Tip**: How to do something a bit more efficiently or effectively.

📝 **Poll**: A zoom poll to help you learn.

🎬 **Demo**: Showing off something more advanced so you know what you can use R for in the future

Before we get started, lets load in the `tidyverse` package that we used in Part 3. We will use this today for subsetting our data and for making plots.

```{r}
# if you haven't installed tidyverse before, uncomment this line 
#install.packages("tidyverse")

# if you have previously installed tidyverse, load it into the current R session
library(tidyverse)
```

### 🥊 Challenge 1: Reading in Data

Load in `data/gapminder.csv` using the `read.csv()` function from previous workshops and assign it to the object **gap**.

```{r}



```

## Summary Statistics Review

In Part 3 we used `summary()`, `table()`, and `mean()` to learn some things about our data. For example:
CLEAN UP

### 🥊 Challenge 2: Summary Statistics

What is the average life expectancy in our data? How much does life expectancy vary?

```{r}

mean(...)

summary(...)

```

We can combine our summary statistics functions with the subsetting functions `filter()` and `select()` from `tidyverse` to ask more specific questions. **For example, what is the average life expectancy in Asia?**

```{r}

gap_asia <- gap %>%
  filter(...)

summary(gap_asia$lifeExp)

```

What if we wanted to print out the average life expectancy for every continent in our data? One option it to write out the code five times but filtering for different continents. 

```{r}

# look at the continents we have
table(gap$continent)
```

```{r}
# show the life expectancy for each continent

# Africa 
gap_africa <- gap %>%
  filter(continent == "Africa")
summary(gap_africa$lifeExp)

# Americas 
gap_america <- gap %>%
  filter(continent == "Americas")
summary(gap_america$lifeExp)

# Asia
gap_asia <- gap %>%
  filter(continent == "Asia")
summary(gap_asia$lifeExp)

# Europe 
gap_europe <- gap %>%
  filter(continent == "Europe")
summary(gap_africa$lifeExp)

# Oceania 
gap_ocean <- gap %>%
  filter(continent == "Oceania")
summary(gap_ocean$lifeExp)


```

🔔 **Question**: What do you think of this approach? Do you notice any problems that it caused?

The code above was time consuming to write out, it was hard to catch errors, and its not easy to edit. What if we wanted to instead show the distribution of GDP per capita? We would have to go back and change five lines of code. 

Instead of copying and pasting code, we can instead put the code into a **function** and call the function whenever we want to show the distribution of life expectancy for a different continent.

### 🎬 **Demo:** Functions instead of copy-pasting code
Here is the same code as above, but placed in a function called `continent_lifeexp`. The function allows us to run the code above for each continent but we do not have to have to code written out five times for each continent. Instead we can called the function `continent_lifeexp`. 

```{r}

continent_lifeexp <- function(continent_name){
  # function to return the distribution of life expectancy for all countries in a given continent
  # arguments
  # - continent_name - continent name, string
  
  # limit to a given continent 
  gap_subset <- gap %>%
    filter(continent == continent_name) # note: this is where the value of continent_name will be filled in 
  
  # print out the life expectancy 
  return(summary(gap_subset$lifeExp))
}
  

# we can change the input and it will change the output accordingly 
continent_lifeexp("Asia")

```


## Custom Functions

We have used many functions that are part of base R or `tidyverse`, for example `head()`, `length()`, `summary()` etc. Custom functions are the same but we define the name of the function, what the input is, and what the output is.

The structure of a custom function is:

```{r}
# # structure of code
# function_name <- function(optional arguments){
#   
#   code
# 
# }

# example
print_number <- function(x){
  
  paste0("My number is: ", x)
  
}

print_number(1)
print_number(5)

```

Custom functions in R have four main components:

1.  The `function` keyword: this keyword tells R that we are defining a function.

2.  The **arguments**: These are the inputs to the function that change. In the example, our argument was a variable `x` that can be any number we want. Functions can have multiple arguments and they can be any class of object.

3.  The **code**: this is the code that the function executes each time it is called. In our example, we just have a single line of code pasting together the sentence "My number is" with the argument `x`. The code in a function can be as long as needed.

4.  The **output**: by default, a function returns the last thing that was outputted by the code. It is good practice to add a `return()` statement, in which you put the explicit thing you want to be outputted.

💡 **Tip:** `paste0` concatenates things together into a single string/sentence. It is a helpful function to use inside custom functions.

Here is another example similar to the demo `continent_lifeexp()` where we return the mean life expectancy for a given continent *and* year.

```{r}
mean_lifeexp <- function(continent_name, yr){
  
  # it is good practice to comment what the function does and what its arguments are at the top:
  # function to return life expectancy of continent
  # continent_name = name of continent, string
  # yr = year we want the life expectancy for, numeric
  
  # every time we would have written the continent, we now write 'cont'
  # limit to continent cont
  gap_subset <- gap %>%
    filter(continent == continent_name & year == yr)

  # save the mean life expectancy for that continent
  lifeexp_mean <- gap_subset %>%
    # pull put the life expectancy column into a vector
    pull(lifeExp) %>%
    # take the mean
    mean()
  
  # return the life expectancy
  return(paste0("The average life expectancy in ", continent_name, " in ", yr, " is ", lifeexp_mean))
    
}

```

Note that running this chunk of code does not run the code inside the function! It saves a function called `mean_lifeexp()`, which now appears in our global environment and can be called at any time in our session.

```{r}
# run the function for different continents
mean_lifeexp("Asia", 2007)
mean_lifeexp("Americas", 1952)
```

### 🥊 Challenge 3: Write a Custom Function

Write a function that filters the gap data set to a specific year. Then estimate the average GDP per capita in that year. Write your function such that you can input a different year and the functions outputs the average GDP per capita for that year. Use the code above as a template.

💡 **Tip:** When writing functions, it is helpful to make sure the body of the code does what you want it to without throwing any errors before including it in the function. To do this, first write the code as you normally would outside of a function. Check it works, then replace any values with the arguments that you use in the function!

```{r}






```

💡 **Tip:** Functions are a great tool for improving your code, especially when you find yourself copying and pasting code. To read more about functions we recommend the [Functions chapter in R for Data Science](https://r4ds.hadley.nz/functions.html).

## Plots 
The custom function above allowed us to answer the question, how does life expectancy vary across continents? Another way to ask this question is to make a plot!

### 🎬 **Demo:** Plotting

How does the life expectancy compare across continents? Here's a box plot example:

```{r}

gap %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot(aes(fill = continent)) +
  theme_bw() +
  ylab("Life Expectancy (years)") +
  ggtitle("Life expectancy by continent (1952-2007)")

  
```

**🔔 Question:** What are some things we learn from this plot? Put your answer in the chat.

### ggplot

`ggplot` is a package from `tidyverse` that helps to create highly customizable plots, all of which have the same underlying syntax. 

A `ggplot` is comprised of three fundamental building blocks:

1.  **data**: typically as a data.frame.

2.  **"aes"thetics**: how our variables will be visually represented (x and y axes, colors, shapes etc.)

3.  **geoms**: what type of plot will represent our data (histogram, scatter plot, bar chart etc.)

One thing to remember is that `ggplot` works in **layers**, similar to photoimaging software such as Photoshop. We create a base layer, and then stack layers on top of that base layer. New layers are added to a plot using the addition symbol `+`. 

## Box Plots

Here we re-create the box plot from above.

1.  **data**: The first layer of a `ggplot` is always the data. This creates a canvas on which to add other layers. For us, the data is the `gap` data.frame.

```{r}

ggplot(data = gap)

```

2.  **aesthetics**: Next, we specify which variables we want to visualize and how. Here we want `continent` on the x-axis, `lifeExp` on the y-axis, and `continent` again as the fill color. All aesthetics go inside the `aes()` function. This prepares our canvas.

```{r}

ggplot(data = gap, aes(x = continent, y = lifeExp, fill = continent))
```

3.  **geom**: Finally, we specify what type of plot we actually want. For us, this is a box plot. Note how we add a new layer with the `+` (not the pipe `%>%` or `|>`). Now our canvas has a picture!

```{r}
ggplot(data = gap, aes(x = continent, y = lifeExp, fill = continent)) +
  geom_boxplot()
```

Any other things that we want to add or customize, we can also add as layers. In the following code, we add a title, change the axis titles, and change the theme (the overall look of the plot).

```{r}

ggplot(data = gap, aes(x = continent, y = lifeExp, fill = continent)) +
  geom_boxplot() +
  # add a title
  ggtitle("Life expectancy by continent (1952-2007)") +
  # change the y axis title
  ylab("Life Expectancy (years)") +
  # change the x axis title
  xlab("Continent") +
  # change the overall theme 
  theme_bw() +
  # remove the legend, b/c this is redundant
  theme(legend.position = 'none')
  

```

### 🥊 Challenge 4: What Plots?

Explore the variables in the `gap` data frame. What other questions could we ask from our data? What types of plots could we use to answer these questions?

-   Q: Which continent has the most varied life expectancy? A: Box plot showing distribution of life expectancy by continent.

-   Q: ... A: ...

💡 **Tip:** A helpful way to learn how to create many different types of plots is to look at a [gallery](https://r-graph-gallery.com/index.html) of possible plots in R.

## Scatter Plots

Box plots are helpful when looking at the distribution of a single continuous variable (e.g. life expectancy) and showing that for different groups (e.g. continents). To show the relationship between two continuous variables, a scatter plot is better. For example, **has life expectancy increased with time in Oceania?**

To make a scatter plot, we use the same syntax as above, but rather than `geom_boxplot()` we use `geom_point()`.

```{r}

gap %>%
  # limit data to just counties in oceania
  filter(continent == "Oceania") %>%
  ggplot(aes(x = year, y = lifeExp)) +
  geom_point()

```

🔔 **Question:** Why are there two points in each year?

```{r}
# we can color the points to indicate which country they represent - this is another aesthetic 

gap %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_point()
```

🔔 **Question:** What can we add to the plot to make it more informative and easier to read? Answer in the chat and we'll try out the suggestions.

```{r}

gap %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_point() +
  # add more layers to make the plot more informative
  ...

```

### 🥊 Challenge 5: Create a Scatter Plot

Create a scatter plot that shows the relationship between `lifeExp` and `gdpPercap` . Start with a "simple" plot using `geom_point()`. 

Next, think about edits you could make to the plot that might make it easier to read. For example, changing the x and y axis, adding a title, coloring the points by `continent`, etc.

⚠️ **Warning**: Some things work better than others - for example, grouping the plots by `country` becomes unwieldy because there are so many countries, but grouping by `continent` is easier to read. 

```{r}

gap %>%
  ggplot(...)

```

💡 **Tip:** `ggplot` can be used to create many different plots. [Here](https://r-graph-gallery.com/index.html) are some examples with the underlying code. To learn more about using `ggplot` you can take D-Lab's [R Data Visualization](https://github.com/dlab-berkeley/R-Data-Visualization) workshop!

## 🎬 **Demo:** Combining custom functions with ggplot()
Plotting is often a good area to utilize custom functions because it is common to want to make many plots that change a few things about them. For example, we might want to see how life expectancy has changed over time for each of our countries. Rather than writing this code out for every country, we can make a function. 

```{r}

lifeexp_plot <- function(country_name){
  
  # function plotting life expectancy against over time by country 
  # country_name = name of country, string

  # every time we would have written the continent, we now write cont 
  p <- gap %>%
    filter(country == country_name) %>%
    ggplot(aes(x = year, y = lifeExp)) +
    geom_point() +
    # here we use the paste0() function to allow cont to change in the title
    ggtitle(paste0("Life expectancy over time in ", country_name)) +
    ylab("Life expectancy (years)") +
    xlab("Year") +
    theme_bw()
  
  # return the plot so it prints out when we call the function
  return(p)
  
}

```

Now we have a function called `lifeexp_plot()` in our global environment. We can call it at any time and provide different country names.

```{r}

lifeexp_plot("Germany")
lifeexp_plot("Ecuador")
lifeexp_plot("Korea Rep.")
lifeexp_plot("Korea Dem. Rep.")

```


## Key Points

-     Custom functions are used to make your code easier to write and less error prone - anytime you are copying and pasting chunks of code, consider putting it in a custom function and calling that function when you need to.

-   Using plots to visualize data is a great way to convey information and find patterns in your data.

-   `ggplot` is a great for creating plots.

## Next steps...

Congratulations on finishing D-Lab's R Fundamentals Series! If you would like to continue learning more R, we recommend taking the following workshops, which build on the skills learned in this workshop series.

1.  [R Data Wrangling](https://github.com/dlab-berkeley/R-Data-Wrangling) - learn more `tidyverse()` commands, including how to create summary statistics by groups and to re-shape data.

2.  [R Data Visualization](https://github.com/dlab-berkeley/R-Data-Visualization) - learn more `ggplot()` tools and create a wide array of different data visualizations.