---
title: "R Fundamentals - Part 3 of 4: Data Frames"
output: html_notebook
---

## Learning Objectives

Welcome to Part 3 of the R Fundamentals series. Our goals for today's workshop are:

1.  Use conditional statements to filter rows

2.  Practice generating summary statistics

3.  Learn how to store categorical variables as factors

Throughout this workshop series, we will use the following icons:

🔔 **Question**: A quick question to help you understand what's going on.

🥊 **Challenge**: Interactive exercise. We'll go through these in the workshop!

⚠️ **Warning**: Heads-up about tricky stuff or common mistakes.

💡 **Tip**: How to do something a bit more efficiently or effectively.

🎬 **Demo**: Showing off something more advanced so you know what you can use R for in the future

Before we get started, lets load in the `tidyverse` package that we used in Part 2. We will use this today for subsetting our data and for making plots.

```{r}
# if you haven't install tidyverse before, uncomment this line 
#install.packages("tidyverse")

# if you have previously installed tidyverse, load it into the current R session
library(tidyverse)
```

### 🥊Challenge 1: Reading in data

In Part 2 we learned to read in a csv file and assign it to a data frame called `gap`. Lets do that again. How can you find out what variables we have in our data?

```{r}




```

## Filtering data frames

If we open up the `gap` data frame (click on the gap object in the Global Environment panel), we see a **Filter** button. This allows us to subset the data frame by rows. We can specify that we want to only show rows for Pakistan or only show rows in the year 1992. We can achieve the same thing by writing code.

### Conditional Statements

In order to specify what rows we want to subset, we need to write **conditional statements** that specify which rows we want. Conditional statements check whether something is true or false. To write them, we need to use **logical operators**:

-   `>` bigger than

-   `<` less than

-   `==` equals to (note the double equal signs)

-   `!=` not equal to

-   `%in%` is in

**🛎️ Answer in the poll whether the following statements are TRUE or FALSE?**

```{r}
# check whether the average population is bigger than 5 million 
mean(gap$pop) > 5000000

# check whether the minimum year is less than or equal to 1950?
min(gap$year) <= 1950

# check if there are exactly 6 variables in the gap data
length(names(gap)) == 6
```

### Filtering

We can use the same logical operators to filter (subset) the rows in our data. We will use the function `filter` from `tidyverse`.

```{r}
# look up the documentation for filter
?filter

# return all rows for Pakistan
filter(gap, country == "Pakistan")

# return all rows where the life expectancy is bigger than 80
filter(gap, lifeExp > 80)
```

We can also combine conditional statements using the `&` and `|` (or) operators.

```{r}
# return all rows in Asia with a GSP per capita above or equal to 30K
filter(gap, continent == "Asia" & gdpPercap >= 30000)

# return all rows in Germany or France
filter(gap, country == "Germany" | country == "France")

# return all rows in Mexico but not from the year 2007
filter(gap, country == "Mexico" & year != 2007)
```

### Piping

We can combine our `select` and `filter` commands to subset both columns and rows. An intuitive way to write code that combines multiple functions is to write multiple lines of code that feed into each other. We do this with the pipe operator `%>%` from `tidyverse`. The `%>%` operator takes the output on the left hand side and passes it into the function on the right hand side.

```{r}
# start with the gap data set
gap %>%
  # filter to Mexico - note, we don't need to specify gap inside select() any more
  filter(country == "Mexico") %>%
  # select the columns we want to return
  select(country, year, pop, lifeExp) 

```

We can assign the subset of our original data frame to a new data frame.

```{r}
# save Mexico rows to another data frame
data_mexico <- gap %>%
  # filter to Mexico
  filter(country == "Mexico") %>%
  # select the columns we want to return
  select(country, year, pop, lifeExp) 

dim(data_mexico)
```

💡 If you have R 4.1 or higher, there is now a base R pipe operator `|>` , which works the same as `%>%` but does not need `tidyverse` to have been loaded.

### 🥊 Challenge 2: Combining filter and select

Start with the gap data frame. Use the `%>%` operator and the functions `select()` and `filter`. Subset the data to countries in Asia or Oceania that have a life expectancy above 60, and select the country, year, life expectancy, and GDP per capita variables. Assign your subsetted data to a new data frame called **gap_subset**.

```{r}



```

## Functions

We have already used several functions in this series. For example, we just used `read.csv()` and `head()`. Functions are commands that take an input and return some output. They are identified by the trailing brackets. Inside the brackets go:

1.  **input** - the thing that the function is doing something to

2.  **arguments** - modifications we make to the function

🔔 What is the **input** for the `head()` function?

```{r}
head(gap)

```

⚠️ Not all functions have inputs or arguments - for example, what is the input for the `ls()` function?

```{r}
ls()

```

We can modify functions by changing the arguments in the function. For example, the default number of rows returned by `head()` is 6. We can see that by looking at the `?head` help page. Default values for arguments are shown after the equal sign.

```{r}
# look up the help page for head()
?head

# change the argument n 
head(gap, n = 10)

```

## Summary Statistics

Before creating visualizations, computing basic summary statistics is a good first step after familiarizing yourself with your data. The `summary()` function gives statistics about the distribution of a variable.

```{r}
# summary for life expectancy
summary(gap$lifeExp)

```

**🔔 What do we learn about the countries in our data from this?**

We can also use the `mean()` function to report just the mean of a variable.

```{r}
mean(gap$lifeExp)

```

## Missing Values

### 🥊Challenge 3: Missing values and functions

What do you notice when you use the `mean()` and `summary()` functions on `gap$gdpPercap` ?

```{r}
# try out mean() and summary() on gdpPercap

```

When we have missing values, functions may perform differently than expected. It is important to know if and why there are missing values in our data. In R, missing values are represented with `NA`, which stands for "not available" or "not applicable". We can use the `is.na()` function to count how many missing values there are in a variable.

```{r}
# count the number of missing values in gdpPercap
is.na(gap$gdpPercap) %>% 
  sum()

# we can also get this information from summary()
summary(gap$gdpPercap)

```

**🔔 Do any of the other variables contain missing values? Edit the code below to find out and answer in the poll.**

```{r}
# see what variables we have
names(gap)

# check which ones are missing by changing the variable after $
is.na(gap$year) %>% 
  sum()
```

When we ran `mean(gap$gdpPercap)` it returned `NA` because it can't calculate a mean over missing values. We need to tell it to explicitly ignore the missing values and calculate the mean using the non-missing values.

### 🥊Challenge 4: Looking up function arguments in documentation

Look up the documentation for `mean`. Which argument will let us ignore `NA` values? What value should we set this argument to? Test it out below.

```{r}
# look up the documentation for mean
?mean

# calculate the mean of gdpPercap ignoring missing values 
mean(gap$gdpPercap, ...) # <- edit this

# is this the same as using summary?
summary(gap$gdpPercap)
```

⚠️ Some functions ignore `NA` values by default and some do not. The function documentation will tell you what the default behavior is , or you can find out with trial and error.

### 🥊Challenge 5: Dropping missing values

We might want to drop the missing values from our data entirely. We can combine `filter()` and `is.na()` to remove rows that contain missing values. Try out the code in the poll and choose the option(s) that drop all the rows with missing values for `gdpPercap`.

Reminder: `filter()` takes a conditional statement, checks if it's true, and then **keeps** all rows that the statement is true for. We want to keep all rows where there are **no** missing values for `gdpPercap`.

```{r}
# fill in a conditional statement in filter to keep only rows with non-missing values of gdpPercap
gap_nomissing <- gap %>%
  filter(...)

# check how many rows we've dropped
dim(gap)
dim(gap_nomissing)

# check if we have any missing values for gdpPercap
summary(gap_nomissing$gdpPercap)
```

⚠️When dropping missing values it drops the entire row of data that contained those missing values. This means we loose cells that had non-missing values for other variables. Choosing whether to drop the row entirely or using functions that ignore missing values will depend on what you're trying to do in your data cleaning and analysis.

### **🎬Reporting the missing values for all variables in our data**

The code below uses two functions from **tidyverse**, `summarize()` and `pivot_longer()`, to make a table showing the number of missing values in each variable. We learn these functions in [R Data Wrangling](https://github.com/dlab-berkeley/R-Data-Wrangling), which is a great course to take after R Fundamentals.

```{r}
gap %>%
  summarize(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(),
               names_to = 'variable', 
               values_to = 'NA_count')
```

## Categorical Variables

The functions `summary()` and `mean()` are helpful for **numeric** variables.

```{r}
# two types of numeric variables:

# (1) numeric
class(gap$lifeExp)

# (2) integer 
class(gap$year)
```

What if we try to use them on **character** variables?

```{r}
summary(gap$continent)

summary(gap$country)

```

Instead we can use the `table()` function to print out a table of all the values that `continent` takes and how many times each value appears (it's frequency).

```{r}
table(gap$continent)
```

We can also create a cross-table where we provide two variables to `table()` and it shows the counts for each cross of the two variables.

```{r}
table(gap$continent, gap$year)
```

### Factors

`country` and `continent` are both **categorical** variables, which means they have a fixed and known set of possible values. For example, `continent` can be one of five values: Africa, Americas, Asia, Europe, or Oceania.

⚠️ Not all character variables are categorical: a variable of people's names does not have a limited and known set of values but it is a character variable. Likewise, not all categorical variables are characters: shoe size is a numeric variable but it has a fixed set of values so it is a categorical variable.

In R, **categorical** variables are best stored as **factors** because this tells R that the variable can take one of a fixed set of values. This is helpful when performing calculations or making plots. The set of values that a factor variable can take are called **levels**. For the `continent` variable, the levels are Africa, Americas, Asia, Europe, and Oceania.

To convert a variable to a categorical variable, you use the `factor()` function. By default, it will look for all unique values in the variable and set those as the possible levels.

```{r}
# convert gap$continent into a factor variable
continent_factor <- factor(gap$continent)

# now we can use summary() on it
summary(continent_factor)

```

One benefit of factors is that it stores the potential levels, even if we do not have any values for that level.

```{r}

# create a continent variable where we don't have any countries from Oceania 
continent_missing <- c("Africa", "Africa", "Americas", "Europe", "Asia", "Europe")

# we can convert this to a factor and record all the possible levels
continent_missing_factor <- factor(continent_missing, levels = c("Africa", "Americas", "Asia", "Europe", "Oceania"))

# check the summary() output
summary(continent_missing_factor)

```

We can also use factors to specify the order in which we want the values to appear by ordering the levels. For character factors, the default order is alphabetical. We want to order our continent variable by land size of the continent, from largest to smallest.

```{r}
# provide the levels in the order we want and specify ordered = T
continent_size <- factor(gap$continent, levels = c("Asia", "Africa", "Americas", "Europe", "Oceania"), ordered = T)

# check the summary() output
summary(continent_size)

# is it different to before?
summary(continent_factor)
```

### 🥊Challenge 5: Ordering factors

Create a new variable that is a factor version of `continent` ordered by the population of the continent from smallest to largest. To help you, this code displays the total population for each continent in our data in 2007.

```{r}

gap %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(total_pop = sum(pop)) %>%
  arrange(total_pop)

```

```{r}
# your code here
continent_pop <- ...


```

🔔The code above contains functions we haven't used yet. What do you think `group_by()` and `summarize()` are doing? If we instead wanted to the total population for each `country` in 2007, what should we change?

⚠️ We create factor versions of our `continent` variable outside of the gap data frame. To create new variable or modify existing variables in a data frame we can use the `mutate()` function from **tidyverse**. This is beyond this workshop but is covered in [R Data Wrangling](https://github.com/dlab-berkeley/R-Data-Wrangling).

## 🎬Next time...

Selecting and filtering data are frames are key steps in telling stories from our data. Looking at summary statistics and visualization are the next steps for understanding our data and presenting it to other people.

🔔 Run the chunk of code below. What do you think we are doing?

```{r}
gap %>%
  filter(year == 2007 | year == 1952) %>%
  select(gdpPercap, lifeExp, continent, year) %>%
  ggplot(aes(x = lifeExp, y = log(gdpPercap), color = continent, shape = continent)) +
  geom_point() +
  facet_wrap(~year) +
  ggtitle("Relationship between Life Expectancy and GDP per capita") +
  ylab("GDP per capita (logged)") +
  xlab("Life Expectancy (years)") +
  theme_bw() 

```

## Key Points

-   Use conditional statements with logical operators to filter data frames to a subset of rows
-   The pipe operator `%>%` is a way to pass output from one line of code to the next line of code
-   The functions `summary()` and `table()` produce summary statistics about variables
-   Different functions in R treat missing values differently - always be aware of whether you have missing data
